---
title: "GSM Data Analysis"
author: "Ricky S."
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    mathjax: "default"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 16,
  fig.height = 12,
  fig.dpi = 300,
  fig.align = "center"
)
```

[Source code of this page](https://github.com/ricky1129/SciP/blob/main/GSM.rmd)

We are going to explore through the data analysis process of cells' sc-RNA seq. 

Data source: [NCBI Geo](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE277276)

A copy of the resources, but not the GSM data, used in this document can ba found at [this GitHub repo](https://www.github.com/ricky1129/SciP).

# Step 0: Load all packages needed

```{r Load Package,results='hide',cache=TRUE}
library(dplyr)
library(Seurat)
library(patchwork)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ggplot2)
```

# Step 0.5: Read back a previous dataset (optional)

```{r Read rds, eval=FALSE,cache=TRUE}
GSM_named <- readRDS("~/SciP/GSM_named.rds")
```

# Step 1: Read data and create individual Seurat objects

`data.dir` is the directory to a set of cell data. The directory should be like this: 
```{text}
|- 6A/
| |- barcodes.tsv.gz
| |- features.tsv.gz
| |- matrix.mtx.gz
|- 8A/
| |- ...
```
The following code does a sample to read data. The path need to be changed. 
```{r Read Data Sample,eval=FALSE,cache=TRUE}
A6.data <- Read10X(data.dir = '/Path/to/Data/6A', gene.column = 1)
A6 <- CreateSeuratObject(counts = A6.data, project = "CeliacD", min.cells = 3, min.features = 200)
```

```{r Read Data 6A,include=FALSE,cache=TRUE}
A6.data <- Read10X(data.dir = '~/SciP/GSM/6A', gene.column = 1)
A6 <- CreateSeuratObject(counts = A6.data, project = "CeliacD", min.cells = 3, min.features = 200)
```

Now that we've read the data, we take a look at it: 
```{r SeeA6,cache=TRUE}
A6
```
We can see that 6115 cells are present in this dataset. 

Using similar methods, we read data from other datasets. 

A total of 10 datasets (5 celiac, 5 control) are loaded for this module. 

```{r Read Data 8A, include=FALSE,cache=TRUE}
A8.data <- Read10X(data.dir = '~/SciP/GSM/8A', gene.column = 1)
A8 <- CreateSeuratObject(counts = A8.data, project = "CeliacD", min.cells = 3, min.features = 200)
```
```{r Read Data 3A16A21A22A23A36A38A39A, include=FALSE,cache=TRUE}
A3.data <- Read10X(data.dir = '~/SciP/GSM/3A', gene.column = 1)
A3 <- CreateSeuratObject(counts = A3.data, project = "CeliacD", min.cells = 3, min.features = 200)
A16.data <- Read10X(data.dir = '~/SciP/GSM/16A', gene.column = 1)
A16 <- CreateSeuratObject(counts = A16.data, project = "CeliacD", min.cells = 3, min.features = 200)
A21.data <- Read10X(data.dir = '~/SciP/GSM/21A', gene.column = 1)
A21 <- CreateSeuratObject(counts = A21.data, project = "CeliacD", min.cells = 3, min.features = 200)
A22.data <- Read10X(data.dir = '~/SciP/GSM/22A', gene.column = 1)
A22 <- CreateSeuratObject(counts = A22.data, project = "CeliacD", min.cells = 3, min.features = 200)
A23.data <- Read10X(data.dir = '~/SciP/GSM/23A', gene.column = 1)
A23 <- CreateSeuratObject(counts = A23.data, project = "CeliacD", min.cells = 3, min.features = 200)
A36.data <- Read10X(data.dir = '~/SciP/GSM/36A', gene.column = 1)
A36 <- CreateSeuratObject(counts = A36.data, project = "CeliacD", min.cells = 3, min.features = 200)
A38.data <- Read10X(data.dir = '~/SciP/GSM/38A', gene.column = 1)
A38 <- CreateSeuratObject(counts = A38.data, project = "CeliacD", min.cells = 3, min.features = 200)
A39.data <- Read10X(data.dir = '~/SciP/GSM/39A', gene.column = 1)
A39 <- CreateSeuratObject(counts = A39.data, project = "CeliacD", min.cells = 3, min.features = 200)
```
# Step 2: Assign metadata to the Seurat objects

We need `@meta.data` to identify cells from different conditions and sources. 

In this case, we will assign `'control'` and `'celiac'` as `meta.data$condition`; 
the numeric id will be a good source to identify sources, stored in `meta.data$batch`. 

```{r Metadata 6A,cache=TRUE}
A6@meta.data$condition <- "control"
A6@meta.data$batch <- "6A"
```
```{r Metadata 8A, include=FALSE,cache=TRUE}
A8@meta.data$condition <- "celiac"
A8@meta.data$batch <- "8A"
```

```{r Metadata 3AA16A21A22A23A36A38A39A, include=FALSE,cache=TRUE}
A3@meta.data$condition <- "control"
A3@meta.data$batch <- "3A"
A16@meta.data$condition <- "control"
A16@meta.data$batch <- "16A"
A21@meta.data$condition <- "celiac"
A21@meta.data$batch <- "21A"
A22@meta.data$condition <- "celiac"
A22@meta.data$batch <- "22A"
A23@meta.data$condition <- "celiac"
A23@meta.data$batch <- "23A"
A36@meta.data$condition <- "celiac"
A36@meta.data$batch <- "36A"
A38@meta.data$condition <- "control"
A38@meta.data$batch <- "38A"
A39@meta.data$condition <- "control"
A39@meta.data$batch <- "39A"
```

# Step 3: Merge into a large dataset and then do quality control

We first merge everything we have into a single seurat object. 

```{r Merge,cache=TRUE}
Merge_obj <- merge(A6, y=c(A8,A3,A16,A21,A22,A23,A36,A38,A39), add.cell.ids=c("1","2","3","4","5","6","7","8","9","10"), project="CeliacD")
Merge_obj
```

Joining the layers together:

```{r Join Layers,cache=TRUE}
Merge_obj[['RNA']] <- JoinLayers(Merge_obj[['RNA']])
```

Now genes are in their ID's, not names. To put their names, we need to match the ID into a database, `org.Hs.eg.db`. 

```{r Convert ID,cache=TRUE}
ensembl_ids <- rownames(Merge_obj)
symbols <- mapIds(org.Hs.eg.db, keys = ensembl_ids, column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first") # Map gene names by database
new_names <- ifelse(is.na(symbols), ensembl_ids, symbols) # Delete N/A in the names
new_names <- make.unique(new_names) # Seurat don't accept identical names, so we make unique ones

counts <- Merge_obj[['RNA']]$counts
rownames(counts) <- new_names

GSM_named <- CreateSeuratObject(counts = counts, meta.data = Merge_obj@meta.data)
```

Now we can filter out dead cells by looking at these 3 factors: 
mitochondria precentage, feature RNA and count RNA. To find appropriate boundaries, 
we plot the cells first to see how they distribute. 

```{r Prefilt,cache=TRUE}
GSM_named[["percent.mt"]] <- PercentageFeatureSet(GSM_named, pattern = "^MT-")

# Argument pt.size = 0 is used to remove the points in the violin plot, which can help to visualize the distribution of the data more clearly.
VlnPlot(GSM_named, features = c('nFeature_RNA','nCount_RNA','percent.mt'),ncol=3, pt.size = 0)
```

Apply suitable scales for each factor. This step will delete around 20% of data. 

::: {.callout-note,cache=TRUE}

### Deleted too many? 

Sometimes the quality of datasets are not optimum and some of them contain tons of dead cells. 
In the data used in this document, one of the datasets are polluted, showing more than $100,000$ cells in `GSM_named`
before filtering. However, after filtering, $36,983$ cells are left, which is a reasonable number. 

If you are unsure, use the following rules to help you: a good dataset should aim for 5000 samples. 
Assuming you have $x$ datasets, you should have

$$
5000 \times x \times (100-20) \div 100 = 4000x
$$

samples after filtering. 

:::

```{r filt,cache=TRUE}
GSM_named <- subset(GSM_named, 
                    subset = nFeature_RNA > 200 
                    & nFeature_RNA < 5000 
                    & nCount_RNA < 15000 
                    & nCount_RNA > 1000 
                    & percent.mt < 20)

```



# Step 4: Normalize the data and identify variable features

We first observe the data before normalizing. 

```{r Observe unint,cache=TRUE}
GSM_named
```

## 4.1 Log normalization

Use log normalization. Note that `1e4` and `LogNormalize` are both default arguments and thus do not need to appear in the code. 

```{r Normalize,cache=TRUE}
GSM_named <- NormalizeData(GSM_named, normalization.method = "LogNormalize", scale.factor = 1e4)
```

## 4.2 Highly variable gene selection

We select highly-variable genes in the data to see which can help us identify different types of cells. 

```{r HVG,results='hide',cache=TRUE}
GSM_named <- FindVariableFeatures(GSM_named, selection.method = 'vst', nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(GSM_named), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(GSM_named)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

## 4.3 PCA

First we scale the data to n(0,1):

```{r Scale,cache=TRUE}
GSM_named <- ScaleData(GSM_named)
```

Then we run PCA on this data:

```{r PCA,cache=TRUE}
GSM_named <- RunPCA(GSM_named, features = VariableFeatures(object = GSM_named))
```

We can use elbow plot to view the results:

```{r ELBP,cache=TRUE}
ElbowPlot(GSM_named, ndims = 100)
```

The turning point help us determine the arg `dims` for Clustering. 


## 4.4 UMAP/Clustering

Clustering:

```{r cluster,cache=TRUE}
GSM_named <- FindNeighbors(GSM_named, dims = 1:30)
GSM_named <- FindClusters(GSM_named, resolution = 0.5)

# Look at cluster IDs of the first 5 cells
head(Idents(GSM_named), 5)
```

UMAP:

```{r umap,cache=TRUE}
GSM_named <- RunUMAP(GSM_named, dims = 1:30)
```
Now we use `DimPlot` to see the results:

```{r DimPlot,cache=TRUE}
# DimPlot(GSM_named, reduction = 'umap', label = TRUE)
DimPlot(GSM_named, reduction = 'umap', group.by = 'batch', label = TRUE)
```

# Step 5: Data integration

We need to split the data into batches before we start integration. 

```{r Split,cache=TRUE}
GSM_named[["RNA"]] <- split(GSM_named[["RNA"]], f = GSM_named@meta.data$batch)
GSM_named
```

This data need to be scaled again before we can do PCA on it. 

```{r Scb4pca,cache=TRUE}
GSM_named <- ScaleData(GSM_named) # This time we did not fill the arguments
GSM_named <- RunPCA(GSM_named, features = VariableFeatures(object = GSM_named))
```

## CCA Integration

```{r CCA,cache=TRUE}
# CCA
GSM_named <- IntegrateLayers(
  object = GSM_named, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = 'integrated.cca',
  verbose = FALSE)

# re-join layers after integration
GSM_named[["RNA"]] <- JoinLayers(GSM_named[["RNA"]])

GSM_named <- FindNeighbors(GSM_named, reduction = "integrated.cca", dims = 1:30)
GSM_named <- FindClusters(GSM_named, resolution = 1)
GSM_named <- RunUMAP(GSM_named,dims = 1:30,reduction = 'integrated.cca')

DimPlot(GSM_named, reduction = 'umap', label = TRUE)
DimPlot(GSM_named, reduction = 'umap', group.by = 'batch', label = TRUE)
```


# Step 6: Annotation

We now try to match different clusters to its cell types. 
Some of the typically expressed genes are listed below: 

- Epithelium (EPCAM)
- Immune (PTPRC)
- Stroma (DCN, VIM, PECAM1)

- Enterocyte (FABP2, FABP6), 
- Goblet (MUC2), 
- Paneth (DEFA5, DEFA6), 
- TA cell (MKI67, TOP2A), 
- Stem(LGR5, OLFM4, ASCL2). 
- Immune: 
- + T(CD3E), 
- + B(MS4A1), 
- + Plasma (JCHAIN), 
- + Myeloid(AIF1), 
- + Fibroblast (DCN, VIM), 
- + Endothelial (PECAM1)

First we plot the dotplot to see which clusters express which genes.

```{r idkaaa,cache=TRUE}
all_markers <- list(
  "Lineage" = c("EPCAM", "PTPRC", "DCN", "VIM", "PECAM1"),
  "Epithelial" = c(
    "FABP2", "FABP6",
    "MUC2",
    "DEFA5", "DEFA6",
    "MKI67", "TOP2A",
    "LGR5", "OLFM4", "ASCL2"
  ),
  "Immune" = c(
    "CD3E",
    "MS4A1",
    "JCHAIN",
    "AIF1"
  )
)

DotPlot(
  GSM_named,
  features = all_markers,
  cluster.idents = FALSE
) 
```

Read the image and analyse with brain (or just go ask chatgpt), we can then label the clusters. 

```{r label,cache=TRUE}
identity <- c(
  "Plasma Cell",              # 0
  "T Cell",                   # 1
  "T Cell",                   # 2
  "Plasma Cell",              # 3
  "T Cell",                   # 4
  "Enterocyte",               # 5
  "T Cell",                   # 6
  "T Cell",                   # 7
  "Endothelial Cell",         # 8
  "Enterocyte / Plasma Cell", # 9
  "T Cell",                   # 10
  "Endothelial Cell",         # 11
  "Intestinal Stem Cell",     # 12
  "Myeloid Cell",             # 13
  "Plasma Cell / Stromal",    # 14
  "Epithelial Cell",          # 15
  "Fibroblast",               # 16
  "Fibroblast",               # 17
  "B Cell",                   # 18
  "Cycling T Cell",           # 19
  "Endothelial Cell",         # 20
  "Epithelial Cell",          # 21
  "Mesenchymal / Stromal",    # 22
  "Fibroblast",               # 23
  "Fibroblast",               # 24
  "Fibroblast",               # 25
  "Endothelial Cell",         # 26
  "Immune (Unassigned)"       # 27
)
names(identity) <- levels(GSM_named)
GSM_named <- RenameIdents(GSM_named, identity)
```

We can now save the dotplot (with labels) into a file. 
```{r savePlot, results='hide',cache=TRUE}
d <- DotPlot(GSM_named, features = all_markers, cluster.idents = FALSE) 
ggsave("~/SciP/GSM_dotplot.png", d, width = 16, height = 12, dpi = 300, bg = "white")
```

# Step the-very-last: Save the dataset
```{r save,cache=TRUE}
saveRDS(GSM_named, file = "~/SciP/GSM_named.rds")
```